### 2015-04-28 ApTranscriptome_TR.Rmd
Made cartoon figure showing predictions for shifts in transcript expression.

Thermal reactionome of the common ant species *Aphaenogaster picea* and *A. carolinensis*
========================================================================================

@@

To examine the species distribution of BLAST hits in the transcriptome assembly, I used the program [Krona](http://sourceforge.net/p/krona/home/krona/) r citep("doi:10.1186/1471-2105-12-385"). I ...

~~~
KRONA code
~~~

The interactive visualization is available [here]().

@@

**Preliminary [examination](https://minilims1.uvm.edu/BCProject-26-Cahan/methods.html#clustering-of-samples) of the data indicated that the A22_7 and Ar_7 samples may have been switched, so I remove these values from the combined expression data set for the two species.**

@@

```{r remove_contaminants}
# extract transcript names from fasta file rather than loading whole file
system("grep '^>' results/trinity-full/Trinity_cap3_uclust_cont.fa | cut -f 1 -d ' ' | sed -r 's/^.{1}//' > results/trinity-full/cont.list")

@@

```{r temperature_response}
A22.type.table <- table(Ap.response.type[ , 'A22.type'])
Ar.type.table <- table(Ap.response.type[ , 'Ar.type'])

@@

```{r mosaic_plot}
# Reorganize for plotting to show overlap among categories between colonies
type.table <- table(Acar = Ap.response.type[ , 'Ar.type'], Apic = Ap.response.type[ , 'A22.type'])
# reorder
tt2 <- type.table[c("Low", "Intermediate", "High", "Bimodal", "NotResp"), c("Low", "Intermediate", "High", "Bimodal", "NotResp")]

@@

```{r mh_plot}
md <- melt(Ec.cells)

mh_plot <- qplot(x=Apic, y=Acar, data=md, fill=value, geom="tile", ylim = rev(levels(md$Acar))) + 
  scale_fill_gradient2(limits=c(-600, 600)) +
  theme(axis.title = element_text(face="italic")) +
  labs(x = "A. picea", y = "A. carolinensis")

# make subplot of rows that match hypotheses

@@

mh_plot_sub1 <- qplot(x=Apic, y=Acar, data=md_sub1, fill=value, geom="tile", ylim = rev(levels(md_sub1$Acar))) + 
  scale_fill_gradient2(limits=c(-600, 600)) +
  theme(axis.title = element_text(face="italic")) +
  labs(x = "A. picea", y = "A. carolinensis") +
  theme(legend.position = "none") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 3.5, xmax = 4.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

@@

mh_plot_sub2 <- qplot(x=Acar, y=Apic, data=md_sub2, fill=value, geom="tile", ylim = rev(levels(md_sub2$Apic))) + 
  scale_fill_gradient2(limits=c(-600, 600)) +
  theme(axis.title = element_text(face="italic")) +
  labs(x = "A. carolinensis", y = "A. picea") + 
  theme(legend.position = "bottom") + 
  geom_rect(xmin = 2.5, xmax = 3.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 3.5, xmax = 4.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

@@

png("results/matched_observations_Fig1.png")
grid.arrange(mh_plot_sub1, mh_plot_sub2, ncol=1, heights = c(0.45, 0.6))
dev.off()

@@

@@

```{r fig3}
# add labels
gg11 <- gg11 + ggtitle('A') + theme(plot.title = element_text(hjust=0))
gg12 <- gg12 + ggtitle('B') + theme(plot.title = element_text(hjust=0))
# grid.arrange
png(file = "results/genetic_assimilation_Fig3.png")
grid.arrange(gg11, gg12, nrow = 1)
dev.off()
```

@@

png("results/crit_temp_regulation_Fig2.png")
grid.arrange(gg2, gg3, nrow = 2, heights = c(.5,.6))
dev.off()
```

>>>>

---
output: word_document
---
Thermal reactionome of the common ant species *Aphaenogaster picea* and *A. carolinensis*
========================================================================================

@@

To examine the species distribution of BLAST hits in the transcriptome assembly, I used the program [Krona](http://sourceforge.net/p/krona/home/krona/) r citep("doi:10.1186/1471-2105-12-385") which provides a taxonomic profile of all transcripts by BLAST annotation to species.

The interactive visualization is available [here](http://johnstantongeddes.org/assets/files/ApTranscriptome_FA.html).

@@

**Preliminary [examination](https://minilims1.uvm.edu/BCProject-26-Cahan/methods.html#clustering-of-samples) of the data indicated that the A22_7 and Ar_7 samples may have been switched, so I remove these values from the combined expression data set for the two species.**  The 'Ar_7' sample clusters with all the other A22 samples, while the 'A22_7' sample clusters with all the other Ar samples. The parsimonious explanation for this is a labeling mistake. Removing these samples, rather than relabeling, is the conservative approach.

![clustering](results/aphaenogaster_sample_clustering.png)

@@

```{r remove_contaminants}
# extract transcript names from fasta file rather than loading whole file
system("grep '^>' results/trinity-full/Trinity_cap3_uclust_cont.fa | cut -f 1 -d ' ' | sed 's/^.{1}//' > results/trinity-full/cont.list")

@@

```{r temperature_response}
A22.type.table <- table(Ap.response.type$A22.type)
Ar.type.table <- table(Ap.response.type$Ar.type)

@@

```{r mosaic_plot}
# Reorganize for plotting to show overlap among categories between colonies
type.table <- table(Acar = Ap.response.type$Ar.type, Apic = Ap.response.type$A22.type)
# reorder
tt2 <- type.table[c("Low", "Intermediate", "High", "Bimodal", "NotResp"), c("Low", "Intermediate", "High", "Bimodal", "NotResp")]

@@

```{r mh_plot}
md <- reshape2::melt(Ec.dev)

mh_plot <- qplot(x=Apic, y=Acar, data=md, fill=value, geom="tile", ylim = rev(levels(md$Acar))) +
  scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(x = "A. picea", y = "A. carolinensis")

mh_plot
# make subplot of rows that match hypotheses

@@

mh_plot_sub1 <- qplot(x=Apic, y=Acar, data=md_sub1, fill=value, geom="tile", ylim = rev(levels(md_sub1$Acar))) + 
  scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(x = "A. picea", y = "A. carolinensis") +
  theme(legend.position = "none") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 3.5, xmax = 4.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

@@

mh_plot_sub2 <- qplot(x=Acar, y=Apic, data=md_sub2, fill=value, geom="tile", ylim = rev(levels(md_sub2$Apic))) + 
  scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(x = "A. carolinensis", y = "A. picea") + 
  theme(legend.position = "bottom") + 
  geom_rect(xmin = 2.5, xmax = 3.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 3.5, xmax = 4.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

@@

png("results/matched_observations_Fig2.png")
grid.arrange(mh_plot_sub1, mh_plot_sub2, ncol=1, heights = c(0.45, 0.6))
dev.off()

@@

Per suggestion of a reviewer, make a cartoon figure that shows the contrasting results of the *enhance defense* and the *tolerance* hypotheses.

The *enhanced defense hypothesis* posits that temperature adaptation uses existing defense mechanisms, which should result in significant overlap in response. This pattern would be apparent as an excess of transcripts with the same response pattern between species (e.g. more shared "Low" transcripts) and a deficit of transcripts that shift to other response patterns or become not responsive.

```{r ed_cartoon}
enhanced_defense_dat <- md_sub1
enhanced_defense_dat$value <- 0
# set non-shared response to random negative value 
enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low"), "value"] <- runif(5, min = -300, max = 0)
enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal"), "value"] <- runif(5, min = -300, max = 0)
# change shared response to positive
enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low" & enhanced_defense_dat$Apic == "Low"), "value"] <- 500
enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal" & enhanced_defense_dat$Apic == "Bimodal"), "value"] <- 300

enhanced_defense_dat

# plot
enhanced_defense_plot <- qplot(x=Apic, y=Acar, data=enhanced_defense_dat, fill=value, geom="tile", ylim = rev(levels(enhanced_defense_dat$Acar))) + 
  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(x = "A. picea", y = "A. carolinensis") +
  theme(legend.position = "none") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 3.5, xmax = 4.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

# title
enhanced_defense_plot <- enhanced_defense_plot + ggtitle('Enhanced defense hypothesis') + theme(plot.title = element_text(hjust=0))
enhanced_defense_plot
```

In contrast, the *tolerance hypothesis* predicts that genes involved in active defense will become non-responsive or shift to other response categories in the better-adapted species.

```{r tol_cartoon}
tolerance_dat <- md_sub1
tolerance_dat$value <- 0
# set non-shared response to random positive value 
tolerance_dat[which(tolerance_dat$Acar == "Low"), "value"] <- runif(5, min = 100, max = 300)
tolerance_dat[which(tolerance_dat$Acar == "Bimodal"), "value"] <- runif(5, min = 100, max = 200)
# change shared response to near zero
tolerance_dat[which(tolerance_dat$Acar == "Low" & tolerance_dat$Apic == "Low"), "value"] <- 30
tolerance_dat[which(tolerance_dat$Acar == "Bimodal" & tolerance_dat$Apic == "Bimodal"), "value"] <- 20

tolerance_dat

# plot
tolerance_plot <- qplot(x=Apic, y=Acar, data=tolerance_dat, fill=value, geom="tile", ylim = rev(levels(tolerance_dat$Acar))) + 
  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(x = "A. picea", y = "A. carolinensis") +
  theme(legend.position = "bottom") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 1.5, ymax = 2.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 3.5, xmax = 4.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

# title
tolerance_plot <- tolerance_plot + ggtitle('Tolerance hypothesis') + theme(plot.title = element_text(hjust=0))
tolerance_plot

# plot together
png("results/predictions_Fig1.png")
grid.arrange(enhanced_defense_plot, tolerance_plot, ncol=1, heights = c(0.45, 0.6))
dev.off()
```

@@

```{r fig3}
# add labels
gg11 <- gg11 + ggtitle('A') + theme(plot.title = element_text(hjust=0))
gg12 <- gg12 + ggtitle('B') + theme(plot.title = element_text(hjust=0))
# grid.arrange
png(file = "results/genetic_assimilation_Fig4.png")
grid.arrange(gg11, gg12, nrow = 1)
dev.off()
```

@@

png("results/crit_temp_regulation_Fig3.png")
grid.arrange(gg2, gg3, nrow = 2, heights = c(.5,.6))
dev.off()
```
